The Effects of Racial Identity and Discrimination on Political Participation and Political Attitudes

### 2024-12-15 draft_finalreport.qmd
updating notebooks

@@

Following the Immigration and Nationality Act of 1965, which allowed for increased immigration into the United States, the United States has become increasingly racially and ethnically diverse. However, studies of political participation and public opinion have often failed to include large samples of racial and ethnic minority Americans, limiting researchers from drawing conclusions on the attitudes of these groups. Using survey data from the 2020 Collaborative Multi-Racial Post-Election Survey, which includes oversamples of racial and ethnic minority populations in the U.S., we investigate the impact of racial identity and racism on political participation and public opinion. First, we analyze whether valuing belonging in a racial group can influence a respondent’s interest in politics. Then, we analyze how personal experiences with racism may influence how respondents value policies aimed at addressing racial and ethnic discrimination. Our findings suggest that racial group belonging is not a predictor of political interest, while gender identity, racial identity, and party affiliation may predict higher levels of political interest. Additionally, gender identity, racial identity, and personal experiences with discrimination may predict increased levels of support for anti-discrimination policies.

@@

## Data

@@

## Variable Selection

@@

## Model Fitting and Evaluation

### Research Question 1: How Racial Group Belonging Impacts Interest in Politics

To model the outcome variable Q29 (interest in participating in politics), we used an ordinal logistic regression with predictors including Q271 (importance of belonging to a racial group to one’s identity), race, age, educational attainment, and political party affiliation. We also included an interaction term between gender and Q271 to assess whether the effect of the importance of belonging to a racial group on one’s identity differs based on one’s gender identity. To assess the model we used a confusion matrix to compare predicted and observed values. Finally, we used the Brant test to check for violations of the proportional odds assumption.

### Research Question 2: How Experiences with Discrimination Affect the Prioritization of Anti-Discrimination Policies

We used a binomial logistic regression to model the outcome variable, Q1R5 (importance placed on anti-discrimination policies), with predictors including race, gender identity, sexual orientation, age, education, community type, experiences with racism, and perception of racism for different groups. An interaction term between gender and race was excluded due to multicollinearity. Stepwise regression and model diagnostics, including VIF and AIC, ensured a stable and interpretable model.

@@

For the first research question, the outcome variable Q29 (interest in participating in politics) was modeled using ordinal logistic regression. Q29, an ordinal variable captured responses to the question: “Some people are very interested in politics while other people can’t stand politics, how about you? Are you…” with response options “Very interested in politics”, “Somewhat interested”, “Not that interested in politics”, and “Not at all interested in politics.” The overall sample reported high levels of interest in politics, with 43% of respondents reporting that they were somewhat interested and one in four (25%) saying they were very interested in politics.

The model also included the predictor variable Q271, a categorical variable that captured responses to the question: “How important is being \[RACIAL GROUP\] to your identity?”, with response options “Extremely important”, “Very important”, “Moderately important”, “Slightly important”, and “Not at all important”. The respondent’s identified racial group was brought into the survey question based on their prior survey responses. Over half (56%) of respondents reported that being a member of their racial group was very or extremely important to their identity. Only 23% of respondents said that being a member of their racial group was slightly or not at all important to their identity.

The model also included multiple predictor variables that control for the impact of other aspects of a respondent’s identity, socioeconomic status, and political preferences These included demographic variables such as the primary race reported by a respondent, their age, and gender, as well as a respondent’s reported level of educational attainment and political party affiliation. Finally, the model included an interaction term between gender and Q271 (the importance a respondent placed in belonging to their racial group) to investigate whether there are combined effects of gender and racial belonging.

(Odds Ratio Plot)

Controlling for age, race, gender identity, educational attainment, and party affiliation, respondents who considered belonging in their racial group to be slightly or moderately important to their identity were marginally more likely to report a higher level of interest in politics than those that considered belonging in their racial group not at all important to their identity, but these effects were not statistically significant. Surprisingly, respondents who considered belonging in their racial group to be very or extremely important to their identity were less likely to report higher levels of interest in politics than the reference category (those who did not place a high importance on belonging to their racial group), holding the other demographic variables constant. These results were statistically significant.

Holding other variables constant, all racial groups in our sample reported increased odds of reporting a higher level of interest in politics compared to white respondents, except for Arab, Middle East & North African (MENA) respondents. However, the effects for respondents of Arab/MENA heritage was not statistically significant. Black respondents were 1.15 times more likely to report a higher level of interest in politics than White respondents. Latino individuals were 1.26 times more likely to report a higher level of interest in politics compared to white individuals. These effects were even more pronounced for smaller racial/ethnic minority groups. Asian Americans were 1.68 times more likely to report a higher level of interest in politics compared to white respondents. Native American (1.49 times more likely) and Native Hawaiian and Pacific Islander respondents (1.78 times more likely) were also more likely to report a higher level of interest in politics compared to white respondents.

For women, the odds of reporting a higher level of interest in politics were 1.58 times higher than men, holding the other demographic variables constant. Non-binary respondents and respondents who specified a different gender identity were less likely to report higher levels of interest in politics than men.  For women and non-binary respondents, reporting that belonging to their racial group was very or extremely important to their racial identity appeared to increase their odds of reporting higher levels of interest in politics, holding other variables constant. However, none of the effects of the interaction term between gender and the importance a respondent placed in belonging to their racial group were statistically significant.

Contrary to the resource model of participation, increases in educational attainment did not result in increased odds of reporting a higher level of interest in politics. Similarly, older respondents reported decreased odds of reporting a higher level of interest in politics compared to the reference category of individuals aged 18-29 years old.

For respondents that identified as Democrats, holding other factors constant, their odds of reporting a higher level of interest in politics decreased 12% compared to Republican respondents. However, these results differed for respondents that identified as Independent or belonging to another party. For Independent respondents, their odds of reporting a higher level of interest in politics were 1.72 times higher than Republican respondents,  holding other variables constant. For respondents that identified with another party, the odds of reporting a higher level of interest in politics were 4.7 times higher than Republican respondents, holding other variables constant. This may indicate that individuals who identify with parties outside of the dominant two party system in the U.S. may already be more interested in politics than typical American respondents.

>>>>

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  message = FALSE,   # Suppress messages
  warning = FALSE    # Suppress warnings
)
```

```{r echo = FALSE}
# Libraries
library(tidyverse)  
library(pROC)
library(vcd)
library(car)
library(MASS) # package for model
library(caret) # package for confusion matrix
library(janitor)
library(brant)
library(knitr)
library(gt)
```

```{r echo = FALSE, results = FALSE}
file_path <- "//Users/pdeguz01/Documents/git/Data/2020 CMPS/DS0001/39096-0001-Data.rda"

load(file_path)          # Load the RDA file
ls()                     # List objects loaded into the environment
dataset <- da39096.0001  # Access the dataset

```

```{r echo = FALSE, results = FALSE}
# Define the required columns
required_cols <- c(
  # Outcome Variable: Stopping discrimination against racial/ethnic minorities
  "Q1R5",

  # Demographic Variables
  "S2_RACE_PRIME",   # Primary Race
  "S3",              # SexualOrientation
  "S3B",             # Gender
  "S5_AGE",          # Age categories
  "S13",             # Highest level of education
  "S14",             # Community lived in

  # Variables capturing experiences with racism
  "Q629R1",          # Unfair treatment due to racial background/ethnicity
  "Q629R2",          # Unfair treatment due to skin color
  "Q629R3",          # Unfair treatment due to gender
  "Q629R4",          # Unfair treatment due to sexuality/orientation
  "Q629R5",          # Unfair treatment due to immigration status
  "Q629R6",          # Unfair treatment due to religion
  "Q629R7",          # Unfair treatment due to accent

  "Q627",            # Treated unfairly or discriminated against because of your race, ethnicity, gender, sexuality, being an immigrant, religious heritage or having an accent?
  "Q633",            # Discrimination impact on life satisfaction

  # Perceptions of discrimination
  "Q619_Q626R1",     # Perceptions of discrimination against Whites
  "Q619_Q626R2",     # Perceptions of discrimination against Blacks
  "Q619_Q626R3",     # Perceptions of discrimination against Asian Americans
  "Q619_Q626R4",     # Perceptions of discrimination against Native Americans
  "Q619_Q626R5",     # Perceptions of discrimination against Immigrants
  "Q619_Q626R6",     # Perceptions of discrimination against Latinos
  "Q619_Q626R7",     # Perceptions of discrimination against GaysAndLesbians
  "Q619_Q626R8"      # Perceptions of discrimination against Muslims
)

```

```{r echo=FALSE, results=FALSE}
# Create a new dataframe with selected variables
dataset_new <- dataset[, required_cols]

# Recode the outcome variable 'Q1R5' for readability
dataset_new$Q1R5 <- factor(
  ifelse(
    dataset_new$Q1R5 == "(0) NO TO: Stopping discrimination against racial/ethnic minorities", 
    "Not important",
    "Important"  # Remaining cases are "Important"
  ),
  levels = c("Not important", "Important") # Define factor levels
)

# Define the columns for perceptions of discrimination
perception_cols <- c(
  "Q619_Q626R1",  # Perceptions of discrimination against Whites
  "Q619_Q626R2",  # Perceptions of discrimination against Blacks
  "Q619_Q626R3",  # Perceptions of discrimination against Asian Americans
  "Q619_Q626R4",  # Perceptions of discrimination against Native Americans
  "Q619_Q626R5",  # Perceptions of discrimination against Immigrants
  "Q619_Q626R6",  # Perceptions of discrimination against Latinos
  "Q619_Q626R7",  # Perceptions of discrimination against GaysAndLesbians
  "Q619_Q626R8",  # Perceptions of discrimination against Muslims
  "Q633"          # Discrimination impact on life satisfaction
)

# Recode 'Don' as 'Don't know' and ensure columns remain factors
dataset_new[perception_cols] <- lapply(dataset_new[perception_cols], function(x) {
  x <- as.character(x)  # Temporarily convert to character for recoding
  x <- ifelse(x == "(5) Don", "(5) Don't know", x)  # Recode 'Don' to 'Don't know'
  factor(x)  # Convert back to factor
})

# Define required columns to convert to factors (if not already defined)
# Assuming `required_cols` is previously defined; otherwise, specify explicitly
dataset_new[required_cols] <- lapply(dataset_new[required_cols], as.factor)

# Verify the structure of the dataset
str(dataset_new[required_cols])

# Display the first few rows to check the changes
head(dataset_new)
```

```{r, echo=FALSE, results=FALSE}
# Data Cleaning and Recoding for dataset_new
dataset_clean <- dataset_new |> 
  mutate(
    # Education recoding
    educ = case_when(
      S13 == "(1) Grades 1-8" ~ "No college experience",
      S13 == "(2) Some High School, but did not graduate" ~ "No college experience",
      S13 == "(3) High School graduate or GED" ~ "No college experience",
      S13 == "(4) Some college" ~ "Some college experience",
      S13 == "(5) Associates, 2-year degree" ~ "College graduate",
      S13 == "(6) Bachelors, 4-year degree" ~ "College graduate",
      S13 == "(7) Post-graduate degree" ~ "College graduate"
    ),
    # Sexual orientation recoding
    sex_orientation = case_when(
      S3 == "(01) Straight, that is, not gay or lesbian" ~ "Straight",
      S3 == "(02) Gay or lesbian" ~ "Queer",
      S3 == "(03) Bisexual" ~ "Queer",
      S3 == "(04) Something else (Specify)" ~ "Queer",
      S3 == "(88) Refused" ~ "Refused/Don't Know",
      S3 == "(99) I don" ~ "Refused/Don't Know"
    ),
    # Urban vs rural recoding
    rural = case_when(
      S14 == "(1) Large urban area" ~ "Urban/suburban",
      S14 == "(2) Large suburb near large city" ~ "Urban/suburban",
      S14 == "(3) Small suburb near small town or city" ~ "Urban/suburban",
      S14 == "(4) Small town or small city" ~ "Urban/suburban",
      S14 == "(5) Rural area" ~ "Rural"
    ),
    # Race recoding
    race_recoded = case_when(
      S2_RACE_PRIME == "(1) White" ~ "White",
      S2_RACE_PRIME == "(2) Hispanic or Latino" ~ "Latino",
      S2_RACE_PRIME == "(3) Black or African American" ~ "Black",
      S2_RACE_PRIME == "(4) Asian American" ~ "Asian American",
      S2_RACE_PRIME == "(5) American Indian/Native American" ~ "AIAN",
      S2_RACE_PRIME == "(6) Arab, Middle Eastern or North African" ~ "Arab/MENA",
      S2_RACE_PRIME == "(7) Native Hawaiian" ~ "NHPI",
      S2_RACE_PRIME == "(8) Not Hawaiian, but other Pacific Islander" ~ "NHPI"
    ),
    # Age recoding
    age_recoded = case_when(
      S5_AGE == "(1) 16-17" ~ "16-17",
      S5_AGE == "(2) 18-29" ~ "18-29",
      S5_AGE == "(3) 30-39" ~ "30-39",
      S5_AGE == "(4) 40-49" ~ "40-49",
      S5_AGE == "(5) 50-59" ~ "50-59",
      S5_AGE == "(6) 60-69" ~ "60-69",
      S5_AGE == "(7) 70 +" ~ "70+"
    )
  ) |>
  mutate(
    # Converting variables to factors
    race_recoded = factor(race_recoded, levels = c("White", "Black", "Asian American", "Latino", "AIAN", "NHPI", "Arab/MENA")),
    age_recoded = factor(age_recoded, levels = c("18-29", "30-39", "40-49", "50-59", "60-69", "70+")),
    educ = factor(educ, levels = c("No college experience", "Some college experience", "College graduate")),
    rural = factor(rural, levels = c("Rural", "Urban/suburban")),
    sex_orientation = factor(sex_orientation, levels = c("Straight", "Queer", "Refused/Don't Know"))
  )

# Take a look at the updated dataset
head(dataset_clean)

```

```{r echo=FALSE, results=FALSE}
# Remove all redundant columns
redundant_columns <- c("Q629R1", "Q629R2", "Q629R3", "Q629R4", 
                       "Q629R5", "Q629R6", "Q629R7",
                       "S2_RACE_PRIME", "S13", "S3", "S5_AGE","S14")

dataset_clean <- dataset_clean[, !(names(dataset_clean) %in% redundant_columns)]
```

```{r echo=FALSE, results=FALSE}
# creating the analysis table for RQ2

names <- c("Intercept",
           "Gender: Women",
           "Gender: Non-Binary",
           "Gender: Something Else",
           "Treated Unfairly Or Discriminated Against: No", # Yes
           "Discrimination Impact On Life Satisfaction: Some", # A lot 
           "Discrimination Impact On Life Satisfaction: A Little",
           "Discrimination Impact On Life Satisfaction: Not At All",
           "Discrimination Impact On Life Satisfaction: Don't Know",
           "Discrimination Impact On Life Satisfaction: Refused",
           "Perceptions Of Discrimination Against Whites: Some", # A lot 
           "Perceptions Of Discrimination Against Whites: A Little",
           "Perceptions Of Discrimination Against Whites: None At All",
           "Perceptions Of Discrimination Against Whites: Don't Know",
           "Perceptions Of Discrimination Against Blacks: Some", # A lot 
           "Perceptions Of Discrimination Against Blacks: A Little",
           "Perceptions Of Discrimination Against Blacks: None At All",
           "Perceptions Of Discrimination Against Blacks: Don't Know",
           "Perceptions Of Discrimination Against Asian Americans: Some", # A lot 
           "Perceptions Of Discrimination Against Asian Americans: A Little",
           "Perceptions Of Discrimination Against Asian Americans: None At All",
           "Perceptions Of Discrimination Against Asian Americans: Don't Know",
           "Perceptions Of Discrimination Against Native Americans: Some", # A lot 
           "Perceptions Of Discrimination Against Native Americans: A Little",
           "Perceptions Of Discrimination Against Native Americans: None At All",
           "Perceptions Of Discrimination Against Native Americans: Don't Know",
           "Perceptions Of Discrimination Against Immigrants: Some", # A lot 
           "Perceptions Of Discrimination Against Immigrants: A Little",
           "Perceptions Of Discrimination Against Immigrants: None At All",
           "Perceptions Of Discrimination Against Immigrants: Don't Know",
           "Perceptions Of Discrimination Against Latinos: Some", # A lot 
           "Perceptions Of Discrimination Against Latinos: A Little",
           "Perceptions Of Discrimination Against Latinos: None At All",
           "Perceptions Of Discrimination Against Latinos: Don't Know",
           "Perceptions Of Discrimination Against Gays And Lesbians: Some", # A lot 
           "Perceptions Of Discrimination Against Gays And Lesbians: A Little",
           "Perceptions Of Discrimination Against Gays And Lesbians: None At All",
           "Perceptions Of Discrimination Against Gays And Lesbians: Don't Know",
           "Perceptions Of Discrimination Against Muslims: Some",# A lot 
           "Perceptions Of Discrimination Against Muslims: A Little",
           "Perceptions Of Discrimination Against Muslims: None At All",
           "Perceptions Of Discrimination Against Muslims: Don't Know",
           "Education: Some College Experience", # No College Experience
           "Education: College Graduate",
           "Orientation: Queer", #Straight
           "Orientation: Refused",
           "Community Type: Urban/Suburban", # Rural
           "Race: Black", # White
           "Race: Asian American",
           "Race: Latino",
           "Race: AIAN",
           "Race: NHPI",
           "Race: Arab/MENA",
           "Age: 30-39 Years", # 18-29
           "Age: 40-49 Years",
           "Age: 50-59 Years",
           "Age: 60-69 Years",
           "Age: 70+ Years"
           )

# Add row names as a new column
analysis_table <- cbind(RowName = names, results_base_model)

# Render the table
table <- analysis_table |>
  gt() |>
  tab_header(
    title = "Odds Ratios with Confidence Intervals and P-Values for base model"
  ) |>
  cols_label(
    RowName = "Description",  # Change the column header for the row names
    OddsRatio = "Odds Ratio",
    LowerCI = "Lower Confidence Interval",
    UpperCI = "Upper Confidence Interval",
    PValue = "P-Value"
  ) |>
  fmt_number(
    columns = c(OddsRatio, LowerCI, UpperCI),
    decimals = 4
  ) |>
  fmt_scientific(
    columns = c(PValue),
    decimals = 2  # Show p-values in scientific notation
  )

# View the table
#table

# Update the column names in the dataframe
colnames(analysis_table) <- c("Description", "Odds Ratio", "Lower Confidence Interval", 
                              "Upper Confidence Interval", "P-Value")

# View analysis table
#analysis_table

# View the updated dataframe
#head(analysis_table)

```

```{r echo=FALSE, results=FALSE}
# Regression model - RQ2
base_model <- glm(ImportanceOfStoppingDiscriminationAgainstMinorities ~ Gender +
               TreatedUnfairlyOrDiscriminatedAgainst + 
               DiscriminationImpactOnLifeSatisfaction +
               PerceptionsOfDiscriminationAgainstWhites + 
               PerceptionsOfDiscriminationAgainstBlacks +
               PerceptionsOfDiscriminationAgainstAsianAmericans + 
               PerceptionsOfDiscriminationAgainstNativeAmericans + 
               PerceptionsOfDiscriminationAgainstImmigrants + 
               PerceptionsOfDiscriminationAgainstLatinos + 
               PerceptionsOfDiscriminationAgainstGaysAndLesbians + 
               PerceptionsOfDiscriminationAgainstMuslims +
               Education + 
                 SexOrientation +
                 Rural +
                 Race +
                 Age,

             data = dataset_clean, family = "binomial")
```

```{r echo=FALSE, results=FALSE}
# Regression Model with interaction term - RQ2
interaction_model <- glm(ImportanceOfStoppingDiscriminationAgainstMinorities ~ 
               TreatedUnfairlyOrDiscriminatedAgainst + 
               DiscriminationImpactOnLifeSatisfaction +
               PerceptionsOfDiscriminationAgainstWhites + 
               PerceptionsOfDiscriminationAgainstBlacks +
               PerceptionsOfDiscriminationAgainstAsianAmericans + 
               PerceptionsOfDiscriminationAgainstNativeAmericans + 
               PerceptionsOfDiscriminationAgainstImmigrants + 
               PerceptionsOfDiscriminationAgainstLatinos + 
               PerceptionsOfDiscriminationAgainstGaysAndLesbians + 
               PerceptionsOfDiscriminationAgainstMuslims +
               Education + 
                 SexOrientation +
                 Rural +
                 Age + 
                 Gender * Race,

             data = dataset_clean, family = "binomial")
```

# ordinal model

# ```{r, echo=FALSE, results=FALSE}
# # data load and cleaning for ordinal regression model
# required_cols_ord <- c("Q29", # Outcome Variable; interest in politics
#                    "Q271", # Predictor Variable; importance of race to identity
#                    "S2_RACE_PRIME", # Race
#                    "S3", # Orientation
#                    "S3B", # Gender
#                    "S5_AGE", # Age Category
#                    "S13", # Educational Attainment
#                    "S14", # Community Type
#                    "S7", # Nativity
#                    "Q21", # Party ID
#                    "S2_M", # Multiracial
#                    
#                    # Religion-related variables
#                    "Q58R1",            # Catholic
#                    "Q58R2",            # Protestant
#                    "Q58R3",            # Other Christian
#                    "Q58R4",            # Jewish
#                    "Q58R5",            # Muslim
#                    "Q58R6",            # Hindu
#                    "Q58R7",            # Buddhist
#                    "Q58R8",            # Mormon/LDS
#                    "Q58R9",            # Folk Religion
#                    "Q58R10",           # Daoist
#                    "Q58R11",           # Shamanism
#                    "Q58R12",           # Ancestral Religion
#                    "Q58R13",           # Spiritual but not religious
#                    "Q58R14",           # Atheist
#                    "Q58R15",            # Agnostic
#                    
#                    # Variables capturing experiences with racism
#                    "Q627", # Treated unfairly or discriminated against ,
#                    "Q628", # Location of discrimination
#                    "Q629R1" # Unfair treatment due to racial background/ethnicity
#                    )
#                  
#                    
#                    
# # Create new df
# data_filtered <- dataset[, required_cols_ord]
# 
# # Convert the required columns to factors
# data_filtered[required_cols] <- lapply(data_filtered[required_cols_ord], as.factor)
# 
# # Convert individual religion columns to numeric
# for (i in 1:15) {
#   col_name <- paste0("Q58R", i, "_numeric")
#   data_filtered[[col_name]] <- as.numeric(sub("^\\((\\d+)\\).*", "\\1", data_filtered[[paste0("Q58R", i)]]))
# }
# 
# # Create a new Religion variable
# data_filtered$Religion <- "None"
# data_filtered$Religion[data_filtered$Q58R1_numeric == 1] <- "Catholic"
# data_filtered$Religion[data_filtered$Q58R2_numeric == 1] <- "Protestant"
# data_filtered$Religion[data_filtered$Q58R3_numeric == 1] <- "Other Christian"
# data_filtered$Religion[data_filtered$Q58R4_numeric == 1] <- "Jewish"
# data_filtered$Religion[data_filtered$Q58R5_numeric == 1] <- "Muslim"
# data_filtered$Religion[data_filtered$Q58R6_numeric == 1] <- "Hindu"
# data_filtered$Religion[data_filtered$Q58R7_numeric == 1] <- "Buddhist"
# data_filtered$Religion[data_filtered$Q58R8_numeric == 1] <- "Mormon/LDS"
# data_filtered$Religion[data_filtered$Q58R9_numeric == 1] <- "Folk Religion"
# data_filtered$Religion[data_filtered$Q58R10_numeric == 1] <- "Daoist"
# data_filtered$Religion[data_filtered$Q58R11_numeric == 1] <- "Shamanism"
# data_filtered$Religion[data_filtered$Q58R12_numeric == 1] <- "Ancestral Religion"
# data_filtered$Religion[data_filtered$Q58R13_numeric == 1] <- "Spiritual but not religious"
# data_filtered$Religion[data_filtered$Q58R14_numeric == 1] <- "Atheist"
# data_filtered$Religion[data_filtered$Q58R15_numeric == 1] <- "Agnostic"
# 
# # Convert Religion to a factor
# data_filtered$Religion <- factor(data_filtered$Religion,
#                                levels = c("Catholic", "Protestant", "Other Christian", "Jewish", "Muslim",
#                                           "Hindu", "Buddhist", "Mormon/LDS", "Folk Religion",
#                                           "Daoist", "Shamanism", "Ancestral Religion",
#                                           "Spiritual but not religious", "Atheist", "Agnostic", "None"))
# 
# # Remove redundant religion columns (Q58R1 to Q58R15 and their numeric versions)
# redundant_cols <- c(
#   "Q58R1", "Q58R2", "Q58R3", "Q58R4", "Q58R5", "Q58R6", "Q58R7", "Q58R8", 
#   "Q58R9", "Q58R10", "Q58R11", "Q58R12", "Q58R13", "Q58R14", "Q58R15",
#   "Q58R1_numeric", "Q58R2_numeric", "Q58R3_numeric", "Q58R4_numeric", 
#   "Q58R5_numeric", "Q58R6_numeric", "Q58R7_numeric", "Q58R8_numeric", 
#   "Q58R9_numeric", "Q58R10_numeric", "Q58R11_numeric", "Q58R12_numeric", 
#   "Q58R13_numeric", "Q58R14_numeric", "Q58R15_numeric"
# )
# 
# # Remove redundant columns from the dataset
# dataset_clean_ord <- data_filtered[, !(names(data_filtered) %in% redundant_cols)]
# 
# dataset_clean_ord <- dataset_clean_ord %>% mutate(educ = case_when(S13 == "(1) Grades 1-8" ~ "No college experience",
#                                                            S13 == "(2) Some High School, but did not graduate" ~ "No college experience",
#                                                            S13 == "(3) High School graduate or GED" ~ "No college experience",
#                                                            S13 == "(4) Some college" ~ "Some college experience",
#                                                            S13 == "(5) Associates, 2-year degree" ~ "College graduate",
#                                                            S13 == "(6) Bachelors, 4-year degree" ~ "College graduate",
#                                                            S13 == "(7) Post-graduate degree" ~ "College graduate"),
#                                           sex_orientation = case_when(S3 == "(01) Straight, that is, not gay or lesbian" ~ "Straight",
#                                                                      S3 == "(02) Gay or lesbian" ~ "Queer",
#                                                                      S3 == "(03) Bisexual" ~ "Queer",
#                                                                      S3 == "(04) Something else (Specify)" ~ "Queer",
#                                                                      S3 == "(88) Refused" ~ "Refused/Don't Know",
#                                                                      S3 == "(99) I don" ~ "Refused/Don't Know"),
#                                           rural = case_when(S14 == "(1) Large urban area" ~ "Urban/suburban",
#                                                             S14 == "(2) Large suburb near large city" ~ "Urban/suburban",
#                                                             S14 == "(3) Small suburb near small town or city" ~ "Urban/suburban",
#                                                             S14 == "(4) Small town or small city" ~ "Urban/suburban",
#                                                             S14 == "(5) Rural area" ~ "Rural"),
#                                           white = case_when(S2_RACE_PRIME == "(1) White" ~ "White", 
#                                                             S2_RACE_PRIME == "(2) Hispanic or Latino" ~ "Non-White",
#                                                             S2_RACE_PRIME == "(3) Black or African American" ~ "Non-White",
#                                                             S2_RACE_PRIME == "(4) Asian American" ~ "Non-White",
#                                                             S2_RACE_PRIME == "(5) American Indian/Native American" ~ "Non-White",
#                                                             S2_RACE_PRIME == "(6) Arab, Middle Eastern or North African" ~ "Non-White",
#                                                             S2_RACE_PRIME == "(7) Native Hawaiian" ~ "Non-White",
#                                                             S2_RACE_PRIME == "(8) Not Hawaiian, but other Pacific Islander" ~ "Non-White"),
#                                           race_recoded = case_when(S2_RACE_PRIME == "(1) White" ~ "White",
#                                                                    S2_RACE_PRIME == "(2) Hispanic or Latino" ~ "Latino",
#                                                                    S2_RACE_PRIME == "(3) Black or African American" ~ "Black",
#                                                                    S2_RACE_PRIME == "(4) Asian American" ~ "Asian American",
#                                                                    S2_RACE_PRIME == "(5) American Indian/Native American" ~ "AIAN",
#                                                                    S2_RACE_PRIME == "(6) Arab, Middle Eastern or North African" ~ "Arab/MENA",
#                                                                    S2_RACE_PRIME == "(7) Native Hawaiian" ~ "NHPI",
#                                                                    S2_RACE_PRIME == "(8) Not Hawaiian, but other Pacific Islander" ~ "NHPI"),
#                                           age_recoded = case_when(S5_AGE == "(1) 16-17" ~ "16-17",
#                                                                   S5_AGE == "(2) 18-29" ~ "18-29",
#                                                                   S5_AGE == "(3) 30-39" ~ "30-39",
#                                                                   S5_AGE == "(4) 40-49" ~ "40-49",
#                                                                   S5_AGE == "(5) 50-59" ~ "50-59",
#                                                                   S5_AGE == "(6) 60-69" ~ "60-69",
#                                                                   S5_AGE == "(7) 70 +" ~ "70+"))
# 
# dataset_clean_ord <- dataset_clean_ord %>% mutate(race_recoded = factor(dataset_clean_ord$race_recoded, levels = c("White", "Black", "Asian American", "Latino", "AIAN", "NHPI", "Arab/MENA")),
#                                          age_recoded = factor(dataset_clean_ord$age_recoded, levels = c("18-29", "30-39", "40-49", "50-59", "60-69", "70+")),
#                                          educ = factor(dataset_clean_ord$educ, levels = c("No college experience", "Some college experience", "College graduate")),
#                                          white = factor(dataset_clean_ord$white, levels = c("White", "Non-White")),
#                                          rural = factor(dataset_clean_ord$rural, levels= c("Rural", "Urban/suburban")),
#                                          sex_orientation = factor(dataset_clean_ord$sex_orientation, levels = c("Straight", "Queer", "Refused/Don't Know")))
# 
# 
# ```

@@

Following the Immigration and Nationality Act of 1965, which allowed for increased immigration into the United States, the United States has become increasingly racially and ethnically diverse. However, studies of political participation and public opinion have often failed to include large samples of racial and ethnic minority Americans, limiting researchers from drawing conclusions on the attitudes of these groups. Using survey data from the 2020 Collaborative Multi-Racial Post-Election Survey, which includes oversamples of racial and ethnic minority populations in the U.S., we investigate the impact of racial identity and racism on political participation and public opinion. First, we analyze whether valuing belonging in a racial group can influence a respondent’s interest in politics. Then, we analyze how personal experiences with racism may influence how respondents value policies aimed at addressing racial and ethnic discrimination. Our findings suggest that racial group belonging is not associated with higher odds  of increased political interest, while gender identity, racial identity, and party affiliation may be associated with higher levels of political interest. Additionally, gender identity, racial identity, and personal experiences with discrimination may be associated with higher odds of increased levels of support for anti-discrimination policies.

@@

@@

@@

To model the outcome variable, interest in participating in politics, in the first research question, we will use an ordinal logistic regression with predictors including the importance of belonging to a racial group to one’s identity, race, age, educational attainment, and political party affiliation. The outcome variable has levels “Not at all interested in politics”, “Not that interested in politics”, “Somewhat interested”, and “Very interested in politics”. We will also include an interaction term between gender and the importance of belonging to a racial group to one’s identity to assess whether the effect of the importance of belonging to a racial group on one’s identity differs based on one’s gender identity. To assess the model we will use a confusion matrix to compare predicted and observed values, and compare the model’s accuracy and no information rate. Finally, we will conduct a Brant test to provide additional evidence regarding whether the proportional odds assumption was violated.

For the second research question, we will make use of a binomial logistic regression to model the outcome variable ’Importance placed on anti-discrimination policies’ with predictors including race, gender identity, sexual orientation, age, education, community type, experiences with racism, and perception of racism for different groups. Additionally, we will include an interaction term between gender and race to explore intersectionality theory and determine which model performs better. After selecting the better model, we will evaluate and assess its performance using various tests such as the confusion matrix, ROC, AUC, accuracy, and other relevant metrics.

@@

For the first research question, the outcome variable, interest in participating in politics, was modeled using ordinal logistic regression. Interest in participating in politics, an ordinal variable, captured responses to the question: “Some people are very interested in politics while other people can’t stand politics, how about you? Are you…” with response options “Very interested in politics”, “Somewhat interested”, “Not that interested in politics”, and “Not at all interested in politics.” The overall sample reported high levels of interest in politics, with 43% of respondents reporting that they were somewhat interested and one in four (25%) saying they were very interested in politics.

| Response to Q29                   | Count  | Share of Total Respondents |
|-----------------------------------|--------|----------------------------|
| Very interested in politics       | 4,306  | 25%                        |
| Somewhat interested               | 7,612  | 43%                        |
| Not that interested in politics   | 3,643  | 21%                        |
| Not at all interested in politics | 1,984  | 11%                        |
| Total                             | 17,545 | 100%                       |

The model also included the predictor variable regarding the importance of belonging to a racial group to one’s identity, a categorical variable that captured responses to the question: “How important is being [RACIAL GROUP] to your identity?”, with response options “Extremely important”, “Very important”, “Moderately important”, “Slightly important”, and “Not at all important”. The respondent’s identified racial group was brought into the survey question based on their prior survey responses. Over half (56%) of respondents reported that being a member of their racial group was very or extremely important to their identity. Only 23% of respondents said that being a member of their racial group was slightly or not at all important to their identity.

The model also included multiple predictor variables that control for the impact of other aspects of a respondent’s identity, socioeconomic status, and political preferences These included demographic variables such as the primary race reported by a respondent, their age, and gender, as well as a respondent’s reported level of educational attainment and political party affiliation. Finally, the model included an interaction term between gender and the importance a respondent placed in belonging to their racial group to investigate whether the relationship between racial belonging and political interest differs depending on gender. This interaction term allows us to evaluate if the effect of racial belonging on political interest is different for males, females, non-binary respondents, or respondents who specified a different gender identity.

(Insert Odds Ratio Plot for RQ1)

Note: Some variable levels are excluded due to filtering for significant variables. To view the reference level for each variable group, please see the below table.

INSERT TABLE

Controlling for age, gender identity, educational attainment, and party affiliation, all racial groups in our sample reported lower odds of reporting a higher level of interest in politics compared to white respondents, except for Arab, Middle East & North African (MENA) respondents. However, the effects for respondents of Arab/MENA heritage was not statistically significant. The odds of Black respondents being in a higher level of interest in politics were 0.87 times as high as White respondents (CI: 0.79-0.95). The odds of Latino respondents being in a higher level of interest in politics were 0.79 times as high as White respondents (CI: 0.72-0.87). These effects were even more pronounced for smaller racial/ethnic minority groups. Asian Americans’ odds of reporting a higher level of interest in politics were 0.60 times as high as that of White respondents (CI: 0.54-0.65).

Increases in educational attainment were associated with higher odds of reporting a higher level of interest in politics. Respondents with some college experience had 1.53 times higher odds of being in a higher level of interest in politics than respondents with no college experience (CI: 1.40-1.66). Respondents who were college graduates had 2.23 times higher odds of being in a higher level of interest in politics than respondents with no college experience (CI: 2.07-2.40). This represents the strongest positive effect in our model.

Respondents in older age groups had odds of being in a higher level of interest in politics that were higher than individuals ages 18-29 years old. Respondents ages 60-69 years old had 1.32 times higher odds of being in a higher level of interest in politics than respondents ages 18-29 years old (CI: 1.21-1.45). Respondents that were 70+ had 1.69 times higher odds of being in a higher level of interest in politics than respondents ages 18-29 years old (CI: 1.51-1.89).

Regarding gender differences, women showed significantly lower odds of political interest compared to men (OR: 0.63, CI: 0.54-0.73). The reference group for the interaction term of the importance of racial belonging and gender identity is male respondents who reported that their belonging to a specific racial group is not at all important to their identity. Respondents who said that their belonging to a specific racial group was “Very important” to their racial identity had 1.19 times higher odds of reporting a higher level of interest in politics than this reference group (CI: 1.03-1.38). Respondents who said that their belonging to a specific racial group was “Extremely important” to their racial identity had 1.64 times higher odds than this reference group (CI: 1.42-1.90).
